Explore & Summarize Data by Suraag Gupta
========================================================

This report is for the Financial Contributions to Presidential Campaigns
dataset. The state used for this dataset is Ohio.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(dplyr)
library(knitr)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
fcpc=read.csv('P00000001-OH.csv')
```

# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots}
str(fcpc)
```

There's almost 170,000 observations in this dataset, with 18 variables. There's
a good chance some of these code blocks will take a fair bit of time to run.

```{r echo=FALSE}
summary(fcpc)
```

Looking at the summary of the variables, some of my observations are that:
- It seems that cmte_id & cand_id are very similar and could possibly have the
same semantic references.
- Looking at contbr_nm, it seems that several people made contributions multiple
times.
- All the contributors are in the state of Ohio.
- The discrepancies between contbr_employer and contbr_occupation are somewhat
strange, as I would expect values such as "RETIRED" to match.
- The contribution quantities are fairly small, looking at the mean and median
values. The large negative number for the minimum is an interesting data point,
I wouldn't expect there to be negative contributions, or "withdrawals" here.
Hopefully, the receipt_desc or memo_text will explain what is going on there.

```{r echo=FALSE}
fcpc$contb_receipt_dt=as.Date(fcpc$contb_receipt_dt, "%e-%b-%y")
```

```{r echo=FALSE}
ggplot(aes(cand_nm, fill=cand_nm), data=fcpc)+
  stat_count(show.legend = FALSE)+coord_flip()
```

This plot shows quite clearly that the primary recipient of contributions in
the state of Ohio was Hilary Clinton, with Bernie Sanders and Trump coming in
second & third, respectively, at nearly half the number. Have to keep in mind
that this isn't for the total amount of contributions, just how many times the
candidate was contributed to.

```{r echo=FALSE}
fcpc_cities=fcpc %>% group_by(contbr_city) %>%
  summarise(count=n(), amount=sum(contb_receipt_amt))

ggplot(aes(contbr_city, count, fill=contbr_city),
       data=subset(fcpc_cities, count>=900))+
  geom_col(show.legend=FALSE)+coord_flip()
```

Looking at some of the observations of contbr_city, it quickly becomes clear
that the data hasn't be cleaned, with several misspellings present for the same
city. Restricting the plot to cities with higher counts shows what most Ohio
residents would expect. Columbus at the top, with Cincinnati close behind.

```{r echo=FALSE, warning=FALSE}
fcpc_contbr=fcpc %>% group_by(contbr_nm) %>%
  summarise(count=n(), amount=sum(contb_receipt_amt))

ggplot(aes(contbr_nm, count), data=subset(fcpc_contbr, count>=100))+
  geom_col()+coord_flip()
```

Somewhat unexpected to see people contributing so many times. The main question
that arises here is who out of these people will appear on the highest amount
plot.

```{r echo=FALSE}
fcpc_occ=fcpc %>% group_by(contbr_occupation) %>%
  summarise(count=n(), amount=sum(contb_receipt_amt))

ggplot(aes(contbr_occupation, count, fill=contbr_occupation),
       data=subset(fcpc_occ, count>=500))+
  geom_col(show.legend=FALSE)+coord_flip()
```

I wasn't expecting to see the Retired contributors make up such a significant
portion of the dataset, this should be interesting to investigate soon.

```{r echo=FALSE, warning=FALSE}
ggplot(aes(log10(contb_receipt_amt)), data=fcpc)+
  geom_histogram(bins=50)
```

Here I used a log10 for the contribution amounts here to visualize better. The
majority of donations seem to be between the $10 to $100 range.

```{r echo=FALSE}
ggplot(aes(contb_receipt_dt), data=fcpc)+stat_count()
```

As would be expected here, most of the contributions were made in in the year of
the election, with the majority coming in what appears to be near the middle or
end of the summer.

```{r echo=FALSE}
ggplot(aes(form_tp, fill=form_tp), data=fcpc)+stat_count(show.legend=FALSE)
```

```{r echo=FALSE}
ggplot(aes(election_tp, fill=election_tp), data=fcpc)+
  stat_count(show.legend=FALSE)
```

Most of the contributions are directed towards the general & presidential
elections.

# Univariate Analysis

### What is the structure of your dataset?
The Financial Contributions dataset for Ohio has 167,259 observations of 18
variables. Committee ID, Candidate ID, & Candidate Name seem to refer to the
same pieces of data. Most of the variables are unordered factors, due to being
names, locations, or reference IDs.

### What is/are the main feature(s) of interest in your dataset?
I think the main features of the dataset are the candidates & the contribution
amounts.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
Supporting the above two features would be the contribution dates/the election
type & the occupation of the contributors.
### Did you create any new variables from existing variables in the dataset?
I created new dataframes for the first three plots, so that I could have an
easier time subsetting the data based on counts.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
There were misspellings in the contributor city, zip code, employer, &
occupation features, but I did not attempt to fix these. I changed the
receipt dates from Strings to Dates to make calculations and plots involving
them easier.

# Bivariate Plots Section

```{r echo=FALSE, warning=FALSE, Bivariate_Plots}
ggplot(aes(cand_nm, contb_receipt_amt, fill=cand_nm), data=fcpc)+
  geom_col(show.legend = FALSE)+ylim(0, NA)+coord_flip()
```

In line with the contribution counts, Hillary Clinton maintains the highest
total contribution value, but unexpectedly, John Kasich has a surprisingly
high contribution value ranking at 2nd, while only having the 6th highest
contribution count. This could indicate the types of voters/contributors that
supported him. Similarly, Bernie & Trump switched roles with counts & amounts.

```{r echo=FALSE, warning=FALSE}
ggplot(aes(contb_receipt_dt, cand_nm, color=cand_nm), data=fcpc)+
  geom_point(alpha=0.1, show.legend=FALSE)
```

Here we see that contributions towards the primary elections lasted for a long
while, and consisted of most of the contributions, which is in line with the
earlier plot of election type counts.

```{r echo=FALSE}
ggplot(aes(contbr_city, amount, fill=contbr_city),
       data=subset(fcpc_cities, amount>=100000))+
  geom_col(show.legend=FALSE)+coord_flip()
```

Slightly unexpectedly, while Columbus may have had the highest number of
contributors, Cincinnati had a higher contribution value.

```{r echo=FALSE}
ggplot(aes(contbr_nm, amount), data=subset(fcpc_contbr, amount>=10000))+
  geom_col()+coord_flip()
```

Comparing this plot to that of the one with the highest contribution counts,
there are no matching names. It's possible that contributors will either make
multiple small contributions over a period of time or they will make a few
large contributions.

```{r echo=FALSE, warning=FALSE}
ggplot(aes(contbr_occupation, amount, fill=contbr_occupation),
       data=subset(fcpc_occ, amount>=50000))+
  geom_col(show.legend=FALSE)+ylim(0, NA)+coord_flip()
```

Moving on to the occupations of the contributors, retirees still take the prime
position with donation amounts, that too by a huge margin.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(contb_receipt_dt, contb_receipt_amt), data=fcpc)+
  geom_jitter(alpha=0.1)+geom_smooth()+
  ylim(0, quantile(fcpc$contb_receipt_amt, 0.95))
```

Using the 95% quantile for the contribution amounts and cutting out the
negatives, which after doing some exploration seemed to be mostly refunds, it
seems that most of the contributions came as the election itself approached.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(contb_receipt_dt, contb_receipt_amt),
       data=subset(fcpc, cand_nm=='Clinton, Hillary Rodham'))+
  geom_jitter(alpha=0.1)+geom_smooth()+
  ylim(0, quantile(fcpc$contb_receipt_amt, 0.95))+ggtitle("Hillary Clinton")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(contb_receipt_dt, contb_receipt_amt),
       data=subset(fcpc, cand_nm=='Trump, Donald J.'))+
  geom_jitter(alpha=0.1)+geom_smooth()+
  ylim(0, quantile(fcpc$contb_receipt_amt, 0.95))+ggtitle("Donald Trump")
```

Two interesting things can be seen in this plot. First, most of the
contributions are larger in value, as only a small chunk of the plot has values
less than about $50. Second, contributions to Trump continued even after he won.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(contb_receipt_dt, contb_receipt_amt),
       data=subset(fcpc, cand_nm=='Sanders, Bernard'))+
  geom_jitter(alpha=0.1)+geom_smooth()+
  ylim(0, quantile(fcpc$contb_receipt_amt, 0.95))+ggtitle("Bernie Sanders")
```

The above plots for the top 3 contributed candidates have similar trends. Few
contributions made in the year prior to the election, but generally of higher
value, while the year of the election is where most of the data lies.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(contb_receipt_dt, contb_receipt_amt),
       data=subset(fcpc, cand_nm=='Kasich, John R.'))+
  geom_jitter(alpha=0.1)+geom_smooth()+
  ylim(0, quantile(fcpc$contb_receipt_amt, 0.95))+ggtitle("John Kasich")
```

This plot is fairly different from the previous 3. While the information is
all gathered in a shorter time frame, it seems that, compared to the other
candidates, there are few contributors that donated less than $50, with the
majority of the values seeming to be at $100 or $250.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(contb_receipt_dt, contb_receipt_amt),
       data=subset(fcpc, contbr_occupation=='RETIRED'))+
  geom_jitter(alpha=0.1)+geom_smooth()+
  ylim(0, quantile(fcpc$contb_receipt_amt, 0.95))+ggtitle("Retirees")
```

A similar situation with the individual candidates is visible with the retired
contributors. This trend is likely identical with all candidates and
contributor occupations.

```{r echo=FALSE, warning=FALSE}
ggplot(aes(election_tp, contb_receipt_amt, color=election_tp), data=fcpc)+
  geom_col(show.legend=FALSE)+ylim(0, NA)
```

The contribution amounts based on the election are in line with the
contribution counts, with the majority of it going towards the primary election.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?
Overall, the contribution amounts and contribution counts produced similar
relationships through the dataset, with only a few notable exceptions that were
pointed out above.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
One of the main factors I observed was the date variable. Depending on when the
candidates were no longer considered in the running, contributions either
completely stopped or slowed down significantly. The amount of contributions
increased closer to the election, while the amount of the contribution didn't
seem to be directly affected by the date.

### What was the strongest relationship you found?
It seems that the strongest relationship is between date and amount.

# Multivariate Plots Section

```{r echo=FALSE, warning=FALSE, Multivariate_Plots}
ggplot(aes(contb_receipt_amt, cand_nm), data=fcpc)+
  geom_count(alpha=0.2)+xlim(0, quantile(fcpc$contb_receipt_amt, 0.98))
```

```{r echo=FALSE}
ggplot(aes(contb_receipt_dt, cand_nm, color=election_tp), data=fcpc)+
  geom_count(alpha=0.1)
```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?
The first multivariate plot created reinforced the smaller contribution values
at higher quantities. The second plot illustrated how the funds were split
for particular elections.

### Were there any interesting or surprising interactions between features?
The second plot, with the candidates vs the contribution dates colorized by the
election type was expected, but still somewhat interesting to see. As the
candidates lost in the primary elections, their funds disappeared, and for the
winners, all the funding went directly to the coming general election. It's
likely that this is just a sorting and organizational detail made by the
committee, as the contributors would not need to make the specification as to
which election the funds went toward.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, warning=FALSE, Plot_One}
ggplot(aes(contbr_occupation, count, fill=contbr_occupation),
       data=subset(fcpc_occ, count>=500))+
  geom_col(show.legend=FALSE)+coord_flip()+ylab("Times Contributed")+
  xlab("Contributor Occupation")+ggtitle("Times Contributed by Occupation")
```

### Description One
I thought this plot was interesting because I didn't expect retirees to make up
such a huge portion of the contributors.

### Plot Two
```{r echo=FALSE, warning=FALSE, Plot_Two}
ggplot(aes(cand_nm, contb_receipt_amt, fill=cand_nm), data=fcpc)+
  geom_col(show.legend = FALSE)+ylim(0, NA)+coord_flip()+
  ylab("Total Contributed ($)")+xlab("Candidate")+
  ggtitle("Contribution per Candidate")
```

### Description Two
This plot is interesting because it demonstrates that even with a much larger
amount of financial support, the top 2 candidates here lost the election.
It would be interesting to see if there were similar situations in previous
election years, as Donald Trump has a significant amount of wealth on his own,
which may have also been a factor in his relatively lower amount of
contribution.

### Plot Three
```{r echo=FALSE, warning=FALSE, message=FALSE, Plot_Three}
ggplot(aes(contb_receipt_dt, contb_receipt_amt), data=fcpc)+
  geom_count(alpha=0.075)+geom_smooth()+
  ylim(0, quantile(fcpc$contb_receipt_amt, 0.95))+xlab("Contribution Date")+
  ylab("Contribution Amount ($)")+
  ggtitle("Amount Contributed vs. Date Contributed")
```

### Description Three
Perhaps the most obvious of the three plots, showing that the bulk of the
contributions came in closer to the date of the general election.
------

# Reflection
I was slightly disappointed with what information there was to find in this
dataset. While the insights and plots created certainly had a good amount of
information to share, there weren't as many variables to explore as I would have
liked. In addition, with the sheer quantity of different contributors, cities,
occupations, & employers, it was a little hard to explore everything as much
as I would have liked. There may be more insights to discover using smaller
subsets based on the contributors, cities, and occupations, after cleaning the
data and having a better method to split up the information and plots that rely
on factored variables.